---
title: "Tephritid metabarcoding workshop"
subtitle: "Bioinformatic analysis"
author: "Alexander M. Piper"
date: "`r Sys.Date()`"
output:
  
  html_document:
    highlighter: null
    theme: "flatly"
    code_download: true
    code_folding: show
    toc: true
    toc_float: 
      collapsed: false
      smooth_scroll: true
    df_print: paged    
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# Knitr global setup - change eval to true to run code
library(knitr)
library(tidyverse)
library(kableExtra)
library(plotly)
library(targets)
library(tarchetypes)

knitr::opts_chunk$set(echo = TRUE, eval=FALSE, message=FALSE, warning=FALSE, error=FALSE, fig.show = "hold", fig.keep = "all")
opts_chunk$set(dev = 'png')
source("R/themes.R")
```

# Introduction

In this workshop we will be analysing 5 example samples from a previous batch of samples. 

The morphological identification of these samples recorded the following species:

```{r, eval=TRUE, echo=FALSE}
trap_catches <- tibble::tribble(
  ~Species, ~Trap1, ~Trap6, ~Trap7, ~Trap19, ~Trap20,
 "B. aeroginosa", 0,0,0,1,0,
 "B. alyxiae", 6,1,15,13,0,
 "B. breviaculeus",9,5,18,8,0,
 "B. bryoniae",1,65,35,15,7,
 "B. erubescentis",52,0,0,0,0,
 "B. frauenfeldi",158,164,192,783,0,
 "B. neohumeralis",17,34,20,148,30,
 "B. peninsularis",2,1,1,0,0,
 "B. tenuifascia",1,0,0,0,0,
 "B. tryoni",12,36,31,94,826,
 "D. axanus",3,0,0,0,0,
 "Z. choristus",0,0,0,1,0,
 "Z. strigifinis",1,5,2,0,0
)
trap_catches%>%
  kbl() %>%
  kable_classic("hover", full_width = T)
```

And visually, the trap compositions identified through morphology look like this:

```{r, eval=TRUE, echo=FALSE}
plot <- trap_catches %>%
  pivot_longer(-Species,
               names_to="sample",
               values_to="specimens") %>%
  ggplot(aes(x = sample, y = specimens, fill=Species)) + 
  geom_col() +
  #scale_fill_brewer(palette="Set1")+
  base_theme +
  theme(legend.position = "right") +
  labs(x = "Trap sample",
       y = "Number of morphologically identified specimens") 

ggplotly(plot)
```

# Set up for analysis

## Install and load R packages and setup directories

The first step is to install and load all necessary R packages required for the pipeline, as well as the FASTQC and BLAST+ command line software.

```{r install and load, eval=TRUE, message=FALSE, warning=FALSE, error=FALSE, results=FALSE} 
#Set required packages
.cran_packages <- c(
  "devtools",
  "ggplot2",
  "gridExtra",
  "data.table",
  "tidyverse", 
  "stringdist",
  "patchwork",
  "vegan",
  "seqinr",
  "patchwork",
  "stringi",
  "magrittr",
  "targets",
  "tarchetypes",
  "zen4R",
  "fs"
  )

.bioc_packages <- c(
  "phyloseq",
  "DECIPHER",
  "Biostrings",
  "ShortRead",
  "ggtree",
  "savR",
  "dada2",
  "ngsReports"
  )

.inst <- .cran_packages %in% installed.packages()
if(any(!.inst)) {
   install.packages(.cran_packages[!.inst])
}
.inst <- .bioc_packages %in% installed.packages()
if(any(!.inst)) {
  if (!requireNamespace("BiocManager", quietly = TRUE)){
    install.packages("BiocManager")
  }
  BiocManager::install(.bioc_packages[!.inst], ask = F)
}

#Load all published packages
sapply(c(.cran_packages,.bioc_packages), require, character.only = TRUE)

# Install and load github packages
devtools::install_github("alexpiper/seqateurs", dependencies = TRUE)
library(seqateurs)

devtools::install_github("alexpiper/taxreturn", dependencies = TRUE)
library(taxreturn)

devtools::install_github("alexpiper/afdscraper", dependencies = TRUE)
library(afdscraper)

devtools::install_github("mikemc/speedyseq", dependencies = TRUE)
library(speedyseq)

#Install bbmap if its not in $path or in bin folder
if(Sys.which("bbduk") == "" & !file.exists("bin/bbmap/bbduk.sh")){
  seqateurs::bbmap_install(dest_dir = "bin")
}

#Install fastqc if its not in $path or in bin folder
if(Sys.which("fastqc") == "" & !file.exists("bin/FastQC/fastqc")){
  seqateurs::fastqc_install(dest_dir = "bin")
}

#Install BLAST if its not in $path or in bin folder
if(Sys.which("blastn") == "" & (length(fs::dir_ls("bin", glob="*blastn.exe",recurse = TRUE)) ==0)){
  taxreturn::blast_install(dest_dir = "bin")
}

source("R/dependencies.R")
source("R/functions.R")
source("R/themes.R")
```

## Create directory structure

This step creates the required directory structure for the pipeline to run. 

```{r}
# Create directories
if(!dir.exists("data")){dir.create("data", recursive = TRUE)}
if(!dir.exists("reference")){dir.create("reference", recursive = TRUE)}
if(!dir.exists("output/logs")){dir.create("output/logs", recursive = TRUE)}
if(!dir.exists("output/results")){dir.create("output/results", recursive = TRUE)}
if(!dir.exists("output/rds")){dir.create("output/rds", recursive = TRUE)}
if(!dir.exists("sample_data")){dir.create("sample_data", recursive = TRUE)}
if(!dir.exists("output/results/final")) {dir.create("output/results/final", recursive = TRUE)}
if(!dir.exists("output/results/unfiltered")) {dir.create("output/results/unfiltered", recursive = TRUE)}
if(!dir.exists("output/results/filtered")) {dir.create("output/results/filtered", recursive = TRUE)}
```

## Fetch sequencing reads

Some test sequencing reads have been hosted on Zenodo for this workshop. The below code will download these and put them inside the data folder.

```{r}
# Create directory for data
if(!dir.exists("data/K77JP")) {dir.create("data/K77JP", recursive = TRUE)}
if(!dir.exists("data/K77JP/InterOp")) {dir.create("data/K77JP/InterOp", recursive = TRUE)}

# Download files from zenodo
download_zenodo(
doi = "10.5281/zenodo.7112162",
path = "data/K77JP"
)

# Move the interop files to the interop folder
fs::dir_ls(path="data/K77JP", glob="*.bin") %>%
  purrr::map(function(x){
    fs::file_copy(path = x, new_path = x %>% str_replace("data/K77JP", "data/K77JP/InterOp"))
    file.remove(x)
  })
```

The directory structure should now look something like this:

    root/
    ├── data/
    │   ├── K77JP/
    │     ├── R1.fastq.gz
    │     ├── R2.fastq.gz
    │     ├── runInfo.xml
    │     ├── runParameters.xml
    │     ├── SampleSheet.csv
    │     └── InterOp/
    ├── sample_data/
    ├── reference
    ├── output/
    └── R/

# Fetch reference files

DOWNLOAD FROM ZENODO LINK!


## Create sample sheet 

In order to track samples and relevant QC statistics throughout the metabarcoding pipeline, we will first create a new samplesheet from our input samplesheets. This function requires both the SampleSheet.csv used for the sequencing run, and the runParameters.xml, both of which should have been automatically obtained from the demultiplexed sequencing run folder in the bash step above

```{r create samplesheet, eval=TRUE}
# Find all flowcell subdirectories within the data directory
runs <- dir("data/")
SampleSheet <- list.files(paste0("data/", runs), pattern= "SampleSheet", full.names = TRUE)
runParameters <- list.files(paste0("data/", runs), pattern= "[Rr]unParameters.xml", full.names = TRUE)

# Create samplesheet containing samples and run parameters for all runs
samdf <- create_samplesheet(SampleSheet = SampleSheet, runParameters = runParameters, template = "V4") %>%
  distinct()

# Check if sampleids contain fcid, if not, attatch these
samdf <- samdf %>%
  mutate(sample_id = case_when(
    !str_detect(sample_id, fcid) ~ paste0(fcid,"_",sample_id),
    TRUE ~ sample_id
  ))

# Get a list of the fastq files
fastqFs <- purrr::map(list.dirs("data", recursive=FALSE),
                      list.files, pattern="_R1_", full.names = TRUE) %>%
  unlist() %>%
  str_remove(pattern = "^(.*)\\/") %>%
  str_remove(pattern = "(?:.(?!_S))+$")
fastqFs <- fastqFs[!str_detect(fastqFs, "Undetermined")]

# Find those that are missing in th sample sheet
if (length(setdiff(fastqFs, samdf$sample_id)) > 0) {warning("The fastq file/s: ", setdiff(fastqFs, samdf$sample_id), " are not in the sample sheet") }

# Find those samples in the samplesheet that don't have fastq files
if (length(setdiff(samdf$sample_id, fastqFs)) > 0) {
  warning(paste0("The fastq file: ",
                 setdiff(samdf$sample_id, fastqFs),
                 " is missing, dropping from samplesheet \n")) 
  samdf <- samdf %>%
    filter(!sample_id %in% setdiff(samdf$sample_id, fastqFs))
}

# Add PCR primers to sample sheet
samdf <- samdf %>%
  mutate(
    pcr_primers = "fwhF2-fwhR2nDac;EIF3LminiF4-EIF3lminiR4",
    for_primer_seq = "GGDACWGGWTGAACWGTWTAYCCHCC;GATGCGYCGTTATGCYGATGC",
    rev_primer_seq = "GTRATWGCHCCIGCTAADACHGG;TTRAAYACTTCYARATCRCC"
    )

#Write out sample CSV for use in pipeline
dir.create("sample_data")
write_csv(samdf, "sample_data/Sample_info.csv")
```

The resulting sample data file should look like this:
```{r, eval=TRUE, echo=FALSE}
samdf %>%
  head()%>%
  kable() %>%
  kable_classic("hover", full_width = T)
```


## Create parameters file

The pipeline also requires a locus parameters file which lists important information about the target barcodes, as well as the PHMM model used to clean the loci, as well as the different reference databases that are to be used for taxonomic assignment.

If multiple reference databases are to be used iteratively for assignment, they should be split with a semicolon, and in the order they are to be used for assignment.

The below code generates the loci_params.csv file, however this can also be done in excel.


```{r Create parameters file, eval=TRUE}
params <- tibble(
  pcr_primers = c("fwhF2-fwhR2nDac", "EIF3LminiF4-EIF3lminiR4"),
  target_gene=c("COI", "EIF3L"),
  phmm = c("reference/phmm/Bactrocera_COI.rds", "reference/phmm/Bactrocera_EIF3L.rds"),
  ref_db = c("reference/COI_internal_idtaxa.rds;reference/COI_idtaxa.rds","reference/EIF3L_internal_idtaxa.rds;reference/EIF3L_idtaxa.rds"),
  blast_db = c("reference/COI_internal.fa.gz;reference/COI_hierarchial.fa.gz", "reference/EIF3L_internal.fa.gz;reference/EIF3L_hierarchial.fa.gz"),
  min_length = c(195, 207),
  max_length = c(215, 227),
  genetic_code = c("SGC4", "SGC0"),
  coding = c(TRUE, TRUE),
  target_kingdom = c("Metazoa", "Metazoa"),
  target_phylum = c("Arthropoda","Arthropoda"),
  target_class = c("Insecta", "Insecta"),
  target_order = c("Diptera", "Diptera"),
  target_family = c(NA_character_, NA_character_), 
  target_genus = c(NA_character_, NA_character_),  
  target_species= c(NA_character_, NA_character_),  
  min_reads_per_sample = c(1000, 1000)
)

# Write out the parameters file
write_csv(params, "sample_data/loci_params.csv")
```

The resulting table should look like this:
```{r, eval=TRUE, echo=FALSE}
params %>%
  kable() %>%
  column_spec(3, width = "1cm")%>%
  column_spec(4, width = "1cm")%>%
  column_spec(5, width = "1cm") %>%
  kable_classic("hover", full_width = T)
```


# Run the pipeline

Now that the sample data and parameters file is ready, its time to run the pipeline!

## Visualise the pipeline steps

```{r, eval=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# Visualise the planned targets pipeline
tar_glimpse()
```


## Run the targets pipeline
```{r}
tar_make()
```


## Check quality control outputs

There are a few important quality control plots that have been automatically created throughout the pipeline. These should be checked to ensure the sequencing run and analysis has run successfully:

* Sequencing run quality check
  + output/logs/K77JP/PFclusters.pdf
  + output/logs/K77JP/Qscore_L1.pdf
  + output/logs/K77JP/avg_intensity.pdf
* Sample quality check
  + output/logs/K77JP/FASTQC/ngsReports_Fastqc.html
* Index-switching
  + output/logs/K77JP/index_switching.pdf
* Filtering plots
  + output/logs/K77JP/prefilt_qualplots.pdf
  + output/logs/K77JP/postfilt_qualplots.pdf
* DADA2 error model
  + output/logs/K77JP/fcid_errormodel.pdf
* Reads surviving pipeline
  + output/logs/K77JP/read_survival.pdf <span style="color:red">NOT WORKING</span>.


# Analyse the results

The outputs of the pipeline are an ASV table, taxonomy table, and sample data, as well as a phyloseq object containing these 3 components. these are located in the following directories

* output/results/final/seqtab.csv
* output/results/final/taxtab.csv
* output/results/final/samdf.csv
* output/rds/ps_filtered.rds


## Read in data

Here we read in the phyloseq object containing the ASV table, taxonomy table, and sample data
```{r phyloseq, eval=TRUE}
ps <- readRDS("output/rds/ps_filtered.rds")

# Turn the phyloseq object into an R data frame, and apply 0.01% minimum abundance threshold
summary_dat <- ps %>%
  speedyseq::tax_glom(taxrank = "Species") %>% # Merges all ASVs assigned to the same species
  speedyseq::psmelt()  %>% # Transforms to data frame
  filter(Abundance > 0 )  %>%
  dplyr::select(OTU, Sample, Abundance,pcr_primers, sample_id, sample_name, environment, collection_location, collection_date, fcid, rank_names(ps)) %>%
  group_by(sample_id, pcr_primers) %>%
  mutate_at(vars(Abundance), ~ . / sum(.) ) %>%
  filter(Abundance > 1e-4) # Remove all under 0.01% abundance

```


## Heatmaps
```{r heatmap, eval=TRUE}
gg.heatmap <- summary_dat%>%
  mutate(sample_name = factor(sample_name,levels=c("Trap1", "Trap6", "Trap7", "Trap19", "Trap20"))) %>%
  ggplot(aes(x=sample_name, y=Species, fill=Abundance)) +
    geom_tile() +
    scale_fill_viridis_c(labels = scales::percent, na.value = NA, alpha=0.9) +
    scale_y_discrete(limits=rev)+
    base_theme+
    theme(legend.position = "right",
          axis.text.y = element_text(face="italic"),
          axis.title.y = element_blank())+
      labs(x="Sample",
           y="Taxon",
           fill="Relative abundance") +
    facet_grid(~pcr_primers, drop=TRUE)

ggplotly(gg.heatmap)
```

## Barplots
```{r barplot, eval=TRUE}
gg.barplot <- summary_dat%>%
  mutate(sample_name = factor(sample_name,levels=c("Trap1", "Trap6", "Trap7", "Trap19", "Trap20"))) %>%
  ggplot(aes(x=sample_name, y=Abundance, fill=Species)) +
    geom_col(colour="black") +
    #scale_fill_viridis_c(labels = scales::percent, na.value = NA, alpha=0.9) +
    #scale_y_discrete(limits=rev)+
    base_theme+
    theme(legend.position = "none",
          axis.text.y = element_text(face="italic"),
          axis.title.y = element_blank())+
      labs(x="Sample",
           y="Taxon",
           fill="Relative abundance") +
    facet_grid(~pcr_primers, drop=TRUE)

ggplotly(gg.barplot)
```